---
title: "R_Final_Project_Abed_Turaki"
author: "Laila and Nargis"
date: "2024-12-02"
output: pdf_document
toc: true
---
\newpage

\listoffigures

\newpage

# Introduction:

Analyzing Mortality Trends and Influencing Factors in Nigeria

This project focuses on analyzing mortality trends and the influencing factors 
in Nigeria, with a particular emphasis on maternal mortality and its correlation with various socio-economic indicators. Using data spanning several decades, we explore the relationship between public health expenditure and maternal 
mortality rates, the impact of health-related factors such as fertility rates 
and life expectancy, and how economic growth and demographic trends have 
influenced public health outcomes. By applying spatial data exploration 
techniques, time series analysis, and regression models, this project aims to provide a comprehensive understanding of the dynamics driving maternal health outcomes in Nigeria and to highlight areas for potential intervention and improvement.
Our github Repository: https://github.com/lja31/R_Final_Project_Abed_Turaki 



Table: Details about the project

|     Item | Value                               |
|---------:|-------------------------------------|
|  Country | Nigeria                             |
|     Date | 12-12-2024                          |
|   Source | World Bank                          |
|    Study | Health indicators                   |
|   Course | EDE ENVIRON 872L-1                  |
| Students | Nargis Taraki & Laila Abed          |

\newpage


# Data preparation and wrangling

```{r setup, include = TRUE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)    
library(sf)
library(here)
library(mapview) ; mapviewOptions(fgb = FALSE)
library(dplyr)
library(ggplot2)
here()
dataset <- read_csv(here("population_sheet_updated.csv"))
colnames(dataset)

data_long <- dataset %>%
  pivot_longer(
    col='1984':'2023',
    names_to = "Year",
    values_to = "Value"
  )
dataset2 <- pivot_wider(data_long, names_from = 'Indicator_Name', values_from ='Value')  

dataset2$Year <- as.numeric(dataset2$Year)
class(dataset2$Year)

```
\newpage


# Question 1: How did the population growth rate evolve over time in the dataset, and what factors might have contributed to these changes?

```{r, fig.cap="Population of Males and Females Over the Years" }


library(ggplot2)

# Convert Year to numeric for plotting
dataset2$Year <- as.numeric(dataset2$Year)

# Create the plot
ggplot(dataset2, aes(x = Year)) +
  geom_line(aes(y = Population_male, color = "Male"), linewidth = 1) +
  geom_line(aes(y = Population_female, color = "Female"), linewidth = 1) +
  scale_color_manual(values = c("Male" = "blue", "Female" = "pink")) +
  labs(
    title = "Population of Males and Females Over the Years",
    x = "Year",
    y = "Population",
    color = "Gender"
  ) + theme(legend.position = "bottom")

```

```{r, fig.cap="Population of Males and Females Over the Years" }
#Life expectancy 
ggplot(dataset2, aes(x = Year)) +
  geom_line(aes(y = Life_expectancy_atbirth_male, color = "Male"), 
            linewidth = 1) +
  geom_line(aes(y = Life_expectancy_at_birth_female, color = "Female"), 
            linewidth = 1) +
  scale_color_manual(values = c("Male" = "blue", "Female" = "pink")) +
  labs(
    title = "Life Expectancy Of Males and Females Over the Years",
    x = "Year",
    y = "Life Expectancy",
    color = "Gender"
  ) + theme(legend.position = "bottom")
```

```{r, fig.cap="Population of Males and Females Over the Years" }
# Survival to age 65 
ggplot(dataset2, aes(x = Year)) +
  geom_line(aes(y = Survival_to_age_65_male, color = "Male"), linewidth = 1) +
  geom_line(aes(y = Survival_to_age_65_female, color = "Female"), 
            linewidth = 1) +
  scale_color_manual(values = c("Male" = "blue", "Female" = "pink")) +
  labs(
    title = "Survival To Age 65 Males and Females Over the Years",
    x = "Year",
    y = "Survival to Age 65",
    color = "Gender"
  ) + theme(legend.position = "bottom")
```

```{r, fig.cap="Population of Males and Females Over the Years" }
# Mortality rate 
ggplot(dataset2, aes(x = Year)) +
  geom_line(aes(y = Mortality_rate_adult_male, color = "Male"), linewidth = 1) +
  geom_line(aes(y = Mortality_rate_adult_female, color = "Female"), 
            linewidth = 1) +
  scale_color_manual(values = c("Male" = "blue", "Female" = "pink")) +
  labs(
    title = "Mortality Rate Of Adult Males and Females Over the Years",
    x = "Year",
    y = "Mortality Rate",
    color = "Gender"
  ) + theme(legend.position = "bottom")
```

\newpage

## Analysis1:Population Growth Rate Evolution

The population growth in Nigeria shows distinct trends for males and females 
over time. Both male and female populations have been steadily increasing from 
1990 to 2020. The male population (represented by the blue line) consistently remains slightly higher than the female population (depicted by the pink line). 
The growth appears to be exponential, with the rate of increase accelerating 
over time. Key factors contributing to these changes include high fertility 
rates, improved healthcare leading to lower infant mortality, better life expectancy, and possibly changes in migration patterns.
Additionally, we noticed from the graphs that the life expectancy for females is higher than males. However the gap between them is decreasing along the years. 
The Mortality rate for males is higher than females, and the same observation 
the gap is decreasing along the years. Survivals to age 65 years of females are more than males and again the gap is decreasing along the years. 

\newpage

# Question2: What trends were observed in the mortality rate ratio over time,and how might this be indicative of broader health and socioeconomic conditions 
in the country?
```{r, fig.cap="Maternal Mortality Ratio Over Time in Nigeria"  }
# Calculating maternal mortality ratio (MMR) over time in Nigeria
# Load necessary library
library(ggplot2)

# Data Preparation (assuming your dataset is called 'dataset2')
dataset2$Year <- as.numeric(dataset2$Year)

# Filter the data to remove rows with missing values 
dataset2_clean <- dataset2 %>%
  select(Year, Maternal_mortality_ratio) %>%
  filter(!is.na(Maternal_mortality_ratio) & !is.na(Year))

# Plotting Maternal Mortality Ratio Over Time
ggplot(dataset2_clean, aes(x = Year, y = Maternal_mortality_ratio)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +  # Add points to show data points clearly
  labs(title = "Maternal Mortality Ratio Over Time in Nigeria",
       x = "Year",
       y = "Maternal Mortality Ratio") +
  theme_minimal() +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 45, hjust = 1))  

```

\newpage

## Analysis2: Maternal Mortality Ratio (MMR) Trends

The graph of Maternal Mortality Ratio (MMR) over time in Nigeria reveals a 
gradual decrease in MMR from 2000 to 2020. While the decline is slow, it is 
steady. However, despite the decrease, the MMR remains relatively high, above 
1000 per 100,000 live births. This suggests slow improvement in maternal healthcare, with challenges in healthcare infrastructure and access still persisting. Socioeconomic development is also gradual, but there is still significant room for improvement in reducing maternal mortality in Nigeria.


\newpage

# Question3: How do mortality rates relate to  availability of nurses/midwives, prenatal care, and female enrollment in primary education, and what potential factors contribute to these relationships?

```{r, fig.cap="Maternal Mortality Ratio vs. Nurses and Midwives" }
# Load necessary libraries for visualization
library(ggplot2)

# 1. Correlation with Nurses and Midwives
dataset2_clean_nurses <- dataset2 %>%
  select(Year, Maternal_mortality_ratio, Nurses_and_midwives) %>%
  filter(!is.na(Maternal_mortality_ratio) & !is.na(Nurses_and_midwives))

# Calculate and print correlation
cor_nurses <- cor(dataset2_clean_nurses$
Maternal_mortality_ratio, dataset2_clean_nurses$Nurses_and_midwives)
print(paste("Correlation with Nurses and Midwives: ", cor_nurses))

# Visualize Correlation with Nurses and Midwives
ggplot(dataset2_clean_nurses, aes(x = Nurses_and_midwives, 
                                  y = Maternal_mortality_ratio)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Maternal Mortality Ratio vs. Nurses and Midwives",
       x = "Nurses and Midwives",
       y = "Maternal Mortality Ratio")
```

```{r, fig.cap="Maternal Mortality Ratio vs. Pregnant Women Receiving Prenatal Care" }
# 2. Correlation with Pregnant Women Receiving Prenatal Care
dataset2_clean_prenatal <- dataset2 %>%
  select(Year, Maternal_mortality_ratio, 
         Pregnant_women_receiving_prenatal_care) %>%
  filter(!is.na(Maternal_mortality_ratio) & 
           !is.na(Pregnant_women_receiving_prenatal_care))

# Calculate and print correlation
cor_prenatal <- cor(dataset2_clean_prenatal
                    $Maternal_mortality_ratio, dataset2_clean_prenatal$Pregnant_women_receiving_prenatal_care)
print(paste("Correlation with Pregnant Women Receiving Prenatal Care: ", 
            cor_prenatal))

# Visualize Correlation with Pregnant Women Receiving Prenatal Care
ggplot(dataset2_clean_prenatal, aes(x = Pregnant_women_receiving_prenatal_care,
                                    y = Maternal_mortality_ratio)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = 
         "Maternal Mortality Ratio vs. Pregnant Women Receiving Prenatal Care",
       x = "Pregnant Women Receiving Prenatal Care",
       y = "Maternal Mortality Ratio") 
```

```{r, fig.cap="Maternal Mortality Ratio vs. School Enrollment Primary Female" }
# 3. Correlation with School Enrollment Primary Female
dataset2_clean_school <- dataset2 %>%
  select(Year, Maternal_mortality_ratio, School_enrollment_primary_female) %>%
  filter(!is.na(Maternal_mortality_ratio) & 
           !is.na(School_enrollment_primary_female))

# Calculate and print correlation
cor_school <- cor(dataset2_clean_school
                  $Maternal_mortality_ratio, 
                  dataset2_clean_school$School_enrollment_primary_female)
print(paste("Correlation with School Enrollment Primary Female: ", cor_school))

# Visualize Correlation with School Enrollment Primary Female
ggplot(dataset2_clean_school, aes(x = School_enrollment_primary_female,
                                  y = Maternal_mortality_ratio)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Maternal Mortality Ratio vs. School Enrollment Primary Female",
       x = "School Enrollment Primary Female",
       y = "Maternal Mortality Ratio") 

```

\newpage

## Analysis3: Mortality Rates and Related Factors

The analysis of mortality rates and their correlation with various factors shows mixed results. The correlation with the number of nurses and midwives is -0.339, which indicates a weak negative correlation. This suggests that as the number of nurses and midwives increases, the maternal mortality ratio slightly decreases. 
In contrast, the correlation with pregnant women receiving prenatal care is 
0.087, indicating a very weak positive correlation and almost no relationship between prenatal care and maternal mortality. The correlation with female 
primary school enrollment is -0.391, showing a moderate negative correlation. 
This suggests that higher female primary school enrollment is associated with 
lower maternal mortality ratios. These relationships suggest that while 
healthcare workforce expansion may have some impact on maternal mortality, other systemic issues, such as access to quality care, play a crucial role. Education, 
on the other hand, appears to have a stronger influence, possibly through better health awareness and decision-making among women.

\newpage

# Question4: How do mortality rates and fertility relate to each other, and what potential factors contribute to this relationship?
```{r, fig.cap="Maternal Mortality Ratio vs. Fertility Rate" }
# 1. Regression with Fertility Rate Total
dataset2_clean_fertility <- dataset2 %>%
  select(Year, Maternal_mortality_ratio, Fertility_rate_total) %>%
  filter(!is.na(Maternal_mortality_ratio) & !is.na(Fertility_rate_total))

# Perform Linear Regression
model_fertility <- lm(Maternal_mortality_ratio ~ 
                        Fertility_rate_total, data = 
                        dataset2_clean_fertility)

# Print Summary of the Model
summary(model_fertility)

# Plotting the regression
ggplot(dataset2_clean_fertility, aes(x = Fertility_rate_total,
                                     y = Maternal_mortality_ratio)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Maternal Mortality Ratio vs. Fertility Rate",
       x = "Fertility Rate Total",
       y = "Maternal Mortality Ratio") 

```

```{r, fig.cap="Relationship Between Maternal Mortality and Women's Life Expectancy" }
# -Regression Analysis: Maternal Mortality Rate vs. Women's Life Expectancy-

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Filter and clean the data
dataset2_clean_life_exp <- dataset2 %>%
  select(Maternal_mortality_ratio, Life_expectancy_at_birth_female) %>%
  filter(!is.na(Maternal_mortality_ratio) &
           !is.na(Life_expectancy_at_birth_female))

# Perform regression analysis
life_exp_regression <- lm(Maternal_mortality_ratio ~ 
            Life_expectancy_at_birth_female, data = dataset2_clean_life_exp)

# Print the summary of the regression model
summary(life_exp_regression)

# Generate a scatterplot with the regression line
ggplot(dataset2_clean_life_exp, aes(x = Life_expectancy_at_birth_female,
                                    y = Maternal_mortality_ratio)) +
  geom_point(color = "blue", size = 2) +  
  geom_smooth(method = "lm", color = "red", se = TRUE) + 
  labs(
    title =
    "Relationship Between Maternal Mortality and Women's Life Expectancy",
    x = "Women's Life Expectancy at Birth (Years)",
    y = "Maternal Mortality Ratio"
  ) + theme(legend.position = "bottom")
```

```{r, fig.cap="Maternal Mortality Ratio vs. Crude Death Rate" }
# 3. Regression with Crude Death Rate
dataset2_clean_crude_death_rate <- dataset2 %>%
  select(Year, Maternal_mortality_ratio, Death_rate_crude) %>%
  filter(!is.na(Maternal_mortality_ratio) & !is.na(Death_rate_crude))

# Perform Linear Regression
model_crude_death_rate <- lm(Maternal_mortality_ratio ~ Death_rate_crude, data
                             = dataset2_clean_crude_death_rate)

# Print Summary of the Model
summary(model_crude_death_rate)

# Plotting the regression
ggplot(dataset2_clean_crude_death_rate, aes(x = Death_rate_crude,
                                            y = Maternal_mortality_ratio)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Maternal Mortality Ratio vs. Crude Death Rate",
       x = "Crude Death Rate",
       y = "Maternal Mortality Ratio") 

```

```{r, fig.cap="Maternal Mortality Ratio vs. Infant Mortality Rate" }
# 4. Regression with Infant Mortality Rate
dataset2_clean_infant_mortality <- dataset2 %>%
  select(Year, Maternal_mortality_ratio, Mortality_rate_infant) %>%
  filter(!is.na(Maternal_mortality_ratio) & !is.na(Mortality_rate_infant))

# Perform Linear Regression
model_infant_mortality <- lm(Maternal_mortality_ratio ~ 
                               Mortality_rate_infant, data = 
                               dataset2_clean_infant_mortality)

# Print Summary of the Model
summary(model_infant_mortality)

# Plotting the regression
ggplot(dataset2_clean_infant_mortality, aes(x = Mortality_rate_infant, y = 
                                              Maternal_mortality_ratio)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Maternal Mortality Ratio vs. Infant Mortality Rate",
       x = "Infant Mortality Rate",
       y = "Maternal Mortality Ratio") 
# -Combined Line Graph: Maternal Mortality Ratio vs. Infant Mortality Rate-

# Filter and clean the data
dataset2_clean_combined <- dataset2 %>%
  select(Year, Maternal_mortality_ratio, Mortality_rate_infant) %>%
  filter(!is.na(Maternal_mortality_ratio) & !is.na(Mortality_rate_infant))

# Generate the combined line graph
ggplot(dataset2_clean_combined, aes(x = Year)) +
  geom_line(aes(y = Maternal_mortality_ratio, color = 
                  "Maternal Mortality Ratio"), size = 1) +
  geom_line(aes(y = Mortality_rate_infant, color = 
                  "Infant Mortality Rate"), size = 1) +
  scale_color_manual(
    name = "Legend",
    values = c("Maternal Mortality Ratio" = 
                 "red", "Infant Mortality Rate" = "blue")
  ) +
  labs(
    title = "Maternal Mortality Ratio and Infant Mortality Rate Over Time",
    x = "Year",
    y = "Rate"
  ) + theme(legend.position = "bottom")


```

\newpage

## Anlaysis 4: Mortality Rates and Fertility Relationship

When examining the relationship between mortality rates and fertility, several regressions were conducted. The regression with fertility rate shows an 
R-squared value of 0.02076, which is very low, and a p-value of 0.5333, 
indicating that there is no significant linear relationship between fertility 
rate and maternal mortality. We though to conduct multi-regression like adding another variable like the age of the mom but we could not find these data. Similarly, the regression with life expectancy at birth yields an R-squared 
value of 0.0003351 and a p-value of 0.9372, suggesting no significant linear relationship between life expectancy and maternal mortality. The regression with the crude death rate also shows an extremely low R-squared value of 0.000001014 
and a p-value of 0.9965, indicating no significant relationship. Lastly, the regression with infant mortality rate shows an R-squared value of 0.003133 and a p-value of 0.8096, also pointing to no significant linear relationship. These results indicate that the factors analyzed do not have a significant linear relationship with maternal mortality. The complex interplay of socioeconomic factors may influence both fertility and mortality in ways not captured by these simple linear regressions, and the possibility of non-linear relationships 
should be considered. Additionally, data quality or reporting issues could 
affect the outcomes of these analyses.

\newpage

# Question5:How does GDP growth over time reflect economic trends in country and what potential factors contribute to these trends?

```{r, fig.cap="Trend Analysis: GDP Growth and Maternal Mortality Ratio" }
# ---- Trend Analysis for GDP Growth and Maternal Mortality Ratio ----

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Ensure data cleaning is done (remove missing values)
dataset2_clean <- dataset2 %>%
  filter(!is.na(GDP_growth) & !is.na(Maternal_mortality_ratio))

# Create time series objects for GDP Growth and Maternal Mortality Ratio
GDP_growth_ts <- ts(dataset2_clean$GDP_growth, start = 
                      c(min(dataset2_clean$Year)), frequency = 1)
MMR_ts <- ts(dataset2_clean$Maternal_mortality_ratio, start
             = c(min(dataset2_clean$Year)), frequency = 1)

# Apply a moving average (MA) for trend smoothing
GDP_growth_trend <- stats::filter(GDP_growth_ts, rep(1/3, 3), sides = 2)
MMR_trend <- stats::filter(MMR_ts, rep(1/3, 3), sides = 2)

# Combine smoothed trends into a data frame for plotting
trend_data <- data.frame(
  Year = dataset2_clean$Year,
  GDP_growth_trend = as.numeric(GDP_growth_trend),
  MMR_trend = as.numeric(MMR_trend)
)

# Plot the trends together
ggplot(trend_data, aes(x = Year)) +
  geom_line(aes(y = GDP_growth_trend, color = 
                  "GDP Growth Trend"), size = 1) +
  geom_line(aes(y = MMR_trend, color = 
                  "Maternal Mortality Ratio Trend"), size = 1) +
  labs(
    title = "Trend Analysis: GDP Growth and Maternal Mortality Ratio",
    x = "Year",
    y = "Smoothed Values",
    color = "Legend"
  ) +
  scale_color_manual(values = c("GDP Growth Trend" = "blue", 
                                "Maternal Mortality Ratio Trend" = "red")) + theme(legend.position = "bottom")



####### Time series
# 1. Data Cleaning: Remove rows with missing values in Maternal Mortality Ratio 
#and calculate differenced values
dataset2_clean_gdp_no_na <- dataset2 %>%
  filter(!is.na(Maternal_mortality_ratio)) %>%  # Keep only rows where Maternal Mortality Ratio is not NA
  mutate(
    GDP_growth_diff = c(NA, diff(GDP_growth)),  # Differenced GDP growth
    MMR_diff = c(NA, diff(Maternal_mortality_ratio))  # Differenced Maternal Mortality Ratio
  )

# Check and handle missing values in differenced GDP growth
sum(is.na(dataset2_clean_gdp_no_na$GDP_growth_diff))  
# Check for missing values in differenced GDP growth

# 2. Remove rows with missing differenced GDP growth values (because they are essential for the analysis)
dataset2_clean_gdp_no_na_no_na <- dataset2_clean_gdp_no_na %>%
  filter(!is.na(GDP_growth_diff))  # Remove rows with missing GDP_growth_diff

# 3. Generate the time series plot for both differenced GDP growth and Maternal Mortality Ratio
ggplot(dataset2_clean_gdp_no_na_no_na, aes(x = Year)) +
  geom_line(aes(y = GDP_growth_diff, color = "Differenced GDP Growth"), 
            size = 1) +
  geom_line(aes(y = MMR_diff, color = "Differenced Maternal Mortality Ratio"), 
            size = 1) +
  geom_point(aes(y = GDP_growth_diff, color = "Differenced GDP Growth"), 
             size = 2, shape = 16) +
  geom_point(aes(y = MMR_diff, color = "Differenced Maternal Mortality Ratio"), size = 2, shape = 16) +
  labs(title = "Differenced Time Series: GDP per Capita Growth and Maternal Mortality Ratio",
       x = "Year", y = "Differenced Value") +
  scale_color_manual(name = "Legend",
                     values = c("Differenced GDP Growth" = "blue", "Differenced Maternal Mortality Ratio" = "red")) +
  theme(legend.position = "bottom")


```

\newpage

## Analysis 5: Economic Trends Reflected in GDP Growth

The analysis of GDP growth over time shows varied trends, reflecting the broader economic shifts in Nigeria. The differenced time series analysis indicates fluctuations in GDP growth, with occasional spikes and drops. A weak positive correlation (correlation coefficient = 0.298) exists between GDP growth and maternal mortality ratio, suggesting that economic improvements alone do not significantly impact maternal health outcomes. Factors such as uneven 
distribution of resources, health expenditure priorities, and persistent poverty likely overshadow the potential benefits of GDP growth. Moreover, the data 
suggest the need for inclusive policies that directly address healthcare infrastructure and maternal care to translate economic growth into tangible 
health outcomes.

\newpage

# Question 6: How does life expectancy at birth differ between males and females and what factors might contribute to these disparities?
```{r, fig.cap="Difference in Life Expectancy at Birth: Female vs Male"}

# Check the summary statistics for male and female life expectancy
summary(dataset2$Life_expectancy_at_birth_female)
summary(dataset2$Life_expectancy_atbirth_male)

#Life Expectancy Difference:
dataset2$Life_expectancy_diff <-
  dataset2$Life_expectancy_at_birth_female - 
  dataset2$Life_expectancy_atbirth_male
summary(dataset2$Life_expectancy_diff)
#
ggplot(dataset2, aes(x = Year)) +
  geom_line(aes(y = Life_expectancy_diff, color = 
                  "Life Expectancy Difference"), size = 1) +
  labs(title = "Difference in Life Expectancy at Birth: Female vs Male",
       x = "Year", y = "Life Expectancy Difference (Female - Male)") +
  scale_color_manual(name = "Legend", values = c("Life Expectancy Difference" 
                                                 = "purple")) 

# Linear regression to explore factors affecting life expectancy difference
lm_model <- lm(Life_expectancy_diff ~ 
                 Maternal_mortality_ratio +
                 Mortality_rate_adult_male 
               + Mortality_rate_adult_female + 
                 Poverty_headcount_ratio + 
                 `Unemployment_male_%ofmalelaborforce`, 
               data = dataset2)
summary(lm_model)


```

\newpage

## Analysis 6: Life Expectancy Disparities Between Genders

The analysis reveals consistent disparities in life expectancy at birth between males and females, with females generally living longer than males. The average life expectancy difference is approximately 1.83 years. Regression analysis suggests that higher maternal mortality rates and adult female mortality rates contribute to the disparity, whereas improvements in male health outcomes and poverty reduction are associated with smaller gaps. Cultural factors, healthcare access, and biological differences likely play roles in this disparity. These results emphasize the importance of targeted interventions to improve male 
health outcomes while continuing to address maternal and female health 
challenges.

\newpage

# Question 7:	How does the immunization rate for measles (children ages 12-23 months) correlate with under-5 mortality rates?
```{r, fig.cap="Immunization Rate for Measles and Under-5 Mortality Rate" }
# Step 1: Extract the relevant columns from the dataset
immunization_measles <- dataset2$`Immunization_measles_children_ages_12-23_m`
under_5_mortality <- dataset2$`Mortality_rate_under-5`

# Step 2: Ensure the columns are numeric
immunization_measles <- as.numeric(immunization_measles)
under_5_mortality <- as.numeric(under_5_mortality)

# Step 3: Remove rows with NA values (missing data)
dataset_clean <- na.omit(data.frame(immunization_measles, under_5_mortality))

# Step 4: Calculate the correlation between immunization rate 
#and under-5 mortality rate
correlation <- cor(dataset_clean$immunization_measles,
                   dataset_clean$under_5_mortality, method = "pearson")

# Print the correlation
print(paste("Correlation between immunization rate 
            and under-5 mortality rate:", correlation))

# Step 5: Visualize the correlation with a scatter plot
library(ggplot2)
ggplot(dataset_clean, aes(x = immunization_measles, y = under_5_mortality)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  geom_point(color = "blue") +
  labs(title = "Immunization Rate for Measles and Under-5 Mortality Rate",
       x = "Immunization Rate for Measles (12-23 months)",
       y = "Under-5 Mortality Rate") + theme(legend.position = "bottom")


```

\newpage

## Analysis 7: Immunization Rate and Under-5 Mortality

There is a moderate negative correlation (-0.519) between the immunization rate 
for measles (children aged 12-23 months) and under-5 mortality rates. This indicates that higher vaccination rates correspond to lower mortality among 
young children. This relationship underscores the critical role of immunization programs in reducing child mortality. However, the correlation also suggests 
that other factors, such as nutrition, sanitation, and broader healthcare 
access, are essential to achieving significant improvements in child survival rates. The scatter plot visualization strengthens the evidence for this inverse relationship, highlighting the importance of continued investment in vaccination efforts.

\newpage

# Question 8: How does the poverty headcount ratio influence  under-5 mortality rate across countries?

```{r, fig.cap="Vaccination Rate vs Under-5 Mortality Rate" }
# Step 1: Extract the poverty rate column
poverty_rate <- dataset2$Poverty_headcount_ratio

# Step 2: Convert to numeric if necessary
poverty_rate <- as.numeric(poverty_rate)

# Step 3: Combine the data into a clean dataset
dataset_clean <- na.omit(data.frame(
  vaccination_rate = immunization_measles,
  poverty_rate = poverty_rate,
  mortality_rate = under_5_mortality
))

# Step 4: Explore the relationship (pairwise correlations)
correlation_vaccine_mortality <- cor(dataset_clean$vaccination_rate, 
                            dataset_clean$mortality_rate, method = "pearson")
correlation_poverty_mortality <- cor(dataset_clean$poverty_rate, 
                            dataset_clean$mortality_rate, method = "pearson")
correlation_vaccine_poverty <- cor(dataset_clean$vaccination_rate, 
                            dataset_clean$poverty_rate, method = "pearson")

# Step 5: Print the correlations
cat("Correlation between Vaccination Rate and Under-5 Mortality Rate:", 
    correlation_vaccine_mortality, "\n")
cat("Correlation between Poverty Rate and Under-5 Mortality Rate:", 
    correlation_poverty_mortality, "\n")
cat("Correlation between Vaccination Rate and Poverty Rate:", 
    correlation_vaccine_poverty, "\n")

# Step 6: Build a multiple linear regression model
model <- lm(mortality_rate ~ vaccination_rate + poverty_rate, data = 
              dataset_clean)

# Step 7: Summarize the model
summary(model)

# Optional Step: Visualize the relationships
library(ggplot2)
ggplot(dataset_clean, aes(x = vaccination_rate, y = mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Vaccination Rate vs Under-5 Mortality Rate")

```

```{r, fig.cap="Poverty Rate vs Under-5 Mortality Rate" }
ggplot(dataset_clean, aes(x = poverty_rate, y = mortality_rate)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Poverty Rate vs Under-5 Mortality Rate")

```

```{r, fig.cap="Relationship Between Poverty Rate and Maternal Mortality Rate" }
# ---- Poverty Rate vs. Maternal Mortality Rate ----

# Filter and clean the data
dataset2_clean_poverty <- dataset2 %>%
  select(Maternal_mortality_ratio, Poverty_headcount_ratio) %>% 
  filter(!is.na(Maternal_mortality_ratio) & !is.na(Poverty_headcount_ratio)) 

# Perform regression analysis
poverty_regression <- lm(Maternal_mortality_ratio ~ Poverty_headcount_ratio, 
                         data = dataset2_clean_poverty)

# Print regression summary
summary(poverty_regression)

# Generate a scatterplot with the regression line
ggplot(dataset2_clean_poverty, aes(x = Poverty_headcount_ratio, y
                                   = Maternal_mortality_ratio)) +
  geom_point(color = "darkblue", size = 2) + 
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Relationship Between Poverty Rate and Maternal Mortality Rate",
    x = "Poverty Headcount Ratio (%)",
    y = "Maternal Mortality Ratio"
  ) 


```

\newpage

## Analysis 8; Poverty Headcount Ratio and Under-5 Mortality

The analysis shows a strong positive correlation (0.782) between poverty 
headcount ratios and under-5 mortality rates, suggesting that higher poverty 
levels are associated with increased child mortality. Conversely, there is a moderate negative correlation (-0.662) between vaccination rates and under-5 mortality rates, demonstrating the mitigating impact of healthcare access. The regression model incorporating poverty and vaccination rates explains about 65%
of the variability in under-5 mortality rates, with poverty showing a stronger influence. These results highlight the pressing need to address poverty and its cascading effects on health while bolstering immunization coverage to protect 
the most vulnerable populations.

\newpage

# question 9: How does public expenditure on health relate to maternal mortality rates over time, and what insights can be drawn from the trends and correlation between these two variables?

```{r, fig.cap="Relationship Between Maternal Mortality Rate and Health Expenditure" }
# ---- Load Required Libraries ----
library(ggplot2)  # For visualizations
library(dplyr)    # For data manipulation

# ---- Data Cleaning and Preprocessing ----

# Step 1: Filter the dataset to keep rows with no missing values in relevant columns
dataset2_clean_health <- dataset2 %>%
  filter(!is.na(Maternal_mortality_ratio) & !is.na(Domestic_general_government_health_expenditure_per_capita))

# Step 2: Replace any missing health expenditure values with the median as a precautionary step
dataset2$Domestic_general_government_health_expenditure_per_capita[
  is.na(dataset2$Domestic_general_government_health_expenditure_per_capita)] <- 
  median(dataset2$Domestic_general_government_health_expenditure_per_capita, 
         na.rm = TRUE)

# Step 3: Check the structure of the dataset (optional step for debugging)
# str(dataset2_clean_health)

# ---- Linear Regression Analysis ----

# Perform regression analysis: Maternal Mortality Rate vs. Health Expenditure
health_expenditure_regression <- lm(
  Maternal_mortality_ratio ~ Domestic_general_government_health_expenditure_per_capita, 
  data = dataset2_clean_health
)

# Print the summary of the regression model
summary(health_expenditure_regression)

# ---- Scatter Plot with Regression Line ----

# Create a scatter plot with a regression line
ggplot(dataset2_clean_health, aes(x = Domestic_general_government_health_expenditure_per_capita, y = Maternal_mortality_ratio)) +
  geom_point(color = "blue", size = 2) +  # Scatterplot points
  geom_smooth(method = "lm", color = "red", se = TRUE) +  # Regression line with confidence interval
  labs(
  title = "Relationship Between Maternal Mortality Rate and Health Expenditure",
    x = "Health Expenditure Per Capita (USD)",
    y = "Maternal Mortality Ratio (Per 100,000 Live Births)"
  ) + theme(legend.position = "bottom")

```

```{r, fig.cap="Trend of Health Expenditure Over Time" }
# ---- Trend Analysis for Health Expenditure ----

# Step 1: Create a time series for health expenditure
health_expenditure_ts <- ts(
dataset2_clean_health$Domestic_general_government_health_expenditure_per_capita,
  start = c(min(dataset2_clean_health$Year)),
  frequency = 1  # Annual data
)

# Step 2: Visualize the trend without decomposition
ggplot(dataset2_clean_health, aes(x = Year, y = Domestic_general_government_health_expenditure_per_capita)) +
  geom_line(color = "blue", size = 1) +  # Line plot for health expenditure
  geom_point(color = "red", size = 2) +  # Highlight individual data points
  labs(
    title = "Trend of Health Expenditure Over Time",
    x = "Year",
    y = "Health Expenditure Per Capita (USD)"
  ) + theme(legend.position = "bottom")


```

\newpage

## Analysis 9:Public expenditure impact on health (maternal mortality) over time
From the plot of trend of health expenditure over time, we noticed Nigeria expenditure for health services increases by time. However some years have significant increase like 2005 the peak and then 2015. In general the government 
is spending more on health services over time (this could be to donors support 
for health programs like HIV mitigation). 
We also noticed that as public expenditure on health services increases, the maternal mortality rate decreases. It makes sense as women receive more care and support the possibility of dying decreases. 

\newpage

# Conclusion
The analysis of mortality trends and influencing factors in Nigeria highlights 
the complexities of health outcomes in a rapidly evolving socio-economic environment. While some progress has been observed, such as the decline in 
maternal mortality and improvements in life expectancy, critical challenges 
remain. High maternal mortality rates, stark gender disparities in life 
expectancy, and the significant impact of poverty on under-5 mortality 
underscore systemic gaps in healthcare access and quality, as well as broader social inequities. Our analysis shows that indicators such as education levels, immunization rates, poverty headcount ratios, and healthcare workforce 
availability significantly contribute to mortality rates in Nigeria, either directly or indirectly.

To drive meaningful change, Nigeria must prioritize investments in healthcare infrastructure, education, and poverty alleviation. Efforts to strengthen immunization programs, improve maternal and child healthcare, and ensure 
equitable distribution of resources are essential to reducing mortality rates further. Integrating economic growth with targeted health and social policies 
will be crucial to overcoming these challenges and fostering sustainable improvements in population health.

Our github Repository: https://github.com/lja31/R_Final_Project_Abed_Turaki 